Component response rate variation drives stability in large complex systems

### 2018-08-02 notebook/ms.Rmd
Reformat for PNAS

---
title: "Component response rate variation drives stability in large complex systems"
author: "A. Bradley Duthie    ( alexander.duthie@stir.ac.uk )"
date: Biological and Environmental Sciences, University of Stirling, Stirling, UK,
  FK9 4LA
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
  html_document: default
  word_document:
    fig_caption: yes
    pandoc_args:
    - --csl
    - science.csl
    reference_docx: docx_template.docx
bibliography: references.bib
header-includes:
- \usepackage{amsmath}
- \usepackage{natbib}
- \usepackage[utf8]{inputenc}
- \bibliographystyle{amnatnat}
linestretch: 1
link-citations: yes
linkcolor: blue
csl: nature.csl
biblio-style: apalike
---

@@

Higher stability in systems with variation in $\gamma$ can be observed by sampling $\gamma_{i}$ values from various distributions. I focus on a uniform distribution where $\gamma \sim \mathcal{U}(0, 2)$ (see Supplementary Information for other distributions, which give similar results). As with the case of $\gamma = \{1.95, 0.5\}$ (Fig. 1b), $E[\gamma] = 1$ when $\gamma \sim \mathcal{U}(0, 2)$, allowing comparison of $\mathbf{M}$ before and after the addition of variation in component response rate. Figure 2 shows a comparison of eigenvalue distributions given $S = 1000$, $C = 0.05$, and $\sigma = 0.4$. As expected [@Tao2010], when $\gamma = 1$, eigenvalues are distributed uniformly in a circle centred at ($-1, 0$) with a radius of $\sigma\sqrt{SC} =$ `r round(sigma*sqrt(S*C), digits = 3)`. Uniform variation in $\gamma$ leads to a non-uniform distribution of eigenvalues[@Ahmadian2015; @Gibbs2017; @Stone2017], some of which are clustered locally near the centre of the distribution, but others of which are spread outside the former radius of `r round(sigma*sqrt(S*C), digits = 3)` (Fig 2b). The clustering and spreading of eigenvalues introduced by $Var(\gamma)$ can destabilise previously stable systems or stabilise systems that are otherwise unstable. But where systems are otherwise too complex to be stable given $\gamma = 1$, the effect of $Var(\gamma)$ can often lead to stability above May's [@May1972; @Allesina2012] threshold $\sigma\sqrt{SC} < 1$.

@@

**Targeted manipulation of $\mathbf{\gamma}$**. To further investigate the potential of $Var(\gamma)$ to be stabilising, I used a genetic algorithm. Genetic algorithms are heuristic tools that mimic evolution by natural selection, and are useful when the space of potential solutions (in this case, possible combinations of $\gamma$ values leading to stability in a large complex system) is too large to search exhaustively[@Hamblin2013]. Generations of selection on $\gamma$ value combinations to minimise $\max\left(\Re(\lambda)\right)$ demonstrated the potential for $Var(\gamma)$ to increase system stability. Across $S = \{2, 3, ..., 39, 40\}$, sets of $\gamma$ values were found that resulted in stable systems with probabilities that were up to four orders of magnitude higher than when $\gamma = 1$ (Fig. 4), meaning that stability could often be achieved by manipulating $S$ $\gamma$ values rather than $S \times S$ $\mathbf{M}$ elements.

**System feasibility given $\mathbf{Var(\gamma)}$** For complex systems in which individual system components represent the density of some tangible quantity, it is relevant to consider the feasibility of the system. Feasibilility assumes that the system is stable and values of all components are positive at equilibrium[@Grilli2017; @Dougoud2018; @Song2018]. This is of particular interest for ecological communities because population density cannot take negative values, meaning that ecological systems need to be feasible for stability to be biologically realistic[@Dougoud2018]. While my results are intended to be general to all complex systems, and not restricted to species networks, I have also performed a feasibility analysis on all matrices $\mathbf{M}$ tested for stability, and additionally for specific types of ecological communities[@Allesina2012] (e.g., competitive, mutualist, predator-prey; see Supplementary Information). Feasibility was unaffected by $Var(\gamma)$, meaning that for pure interacting species networks, variation in component response rate (i.e., species generation time) does not affect stability at biologically realistic species densities.

********************************************************************************

@@

********************************************************************************

Here I have shown that the stability of large systems might often be contigent upon variation in the response rates of their individual components, meaning that factors such as rate of trait evolution (in biological networks), transaction speed (in economic networks), or communication speed (in social networks) need to be considered when investigating the stability of complex systems. Variation in component response rate becomes more likely to be critical for stability as system size increases, and can ultimately increase the overall probability that system stability is observed above that predicted by May's[@May1972] classically derived $\sigma \sqrt{SC}$ criterion. The logic outlined here is general, and potentially applies to any complex system in which individual system components can vary in their reaction rates to system perturbation.

********************************************************************************

@@

********************************************************************************

It is important to emphasise that variation in component response rate is not stabilising per se; that is, adding variation in component response rate to a particular system does not necessarily increase the probability that the system will be stable. Rather, systems that are observed to be stable are more likely to have varying component response rates, and for this variation to be critical to their stability (Fig. 3). This is caused by the shift to a non-uniform distribution of eigenvalues that occurs by introducing $Var(\gamma)$ (Fig. 1b, 2b), which can sometimes cause all of the real components of the eigenvalues of the community matrix $\mathbf{M}$ to become negative, but might also increase the real components of eigenvalues. The mathematics underlying this shift in eigenvalue distribution has been investigated[@Ahmadian2015] and recently applied to questions concerning species density and feasibility[@Gibbs2017; @Stone2017], but has not been interpreted as rates of response of individual system components to perturbation.

********************************************************************************

@@

********************************************************************************

Interestingly, while complex systems were more likely to be stable given variation in component response rate, they were not more likely to be feasible, meaning that stability was not increased when component values were also restricted to being positive at equilibrium. Feasibility is important to consider, particularly for the study of ecological networks of species[@Stone2017; @Grilli2017; @Dougoud2018] because population densities cannot realistically be negative. My results therefore suggest that variation in the rate of population responses to perturbation (e.g., due to differences in generation time among species) is unlikely to be critical to the stability of purely multi-species interaction networks (see also Supplementary Information). Nevertheless, ecological interactions do not exist in isolation in empirical systems[@Patel2018], but instead interact with evolutionary, abiotic, or social-economic systems. The relevance of component response rate for complex system stability should therefore not be ignored in the broader context of ecological communities.

********************************************************************************

@@

********************************************************************************

In the above, $f_{i}$ and $g_{j}$ are functions that define the effects of all species densities and trait values on the density of a species $i$ and the value of trait $j$, respectively. Patel et al. were interested in stability when the evolution of traits was relatively slow or fast in comparison with the change in species densities[@Patel2018], and this is modulated in the above by the scalar $\epsilon$. The value of $\epsilon$ thereby determines the timescale separation between ecology and evolution, with high $\epsilon$ modelling relatively fast evolution and low $\epsilon$ modelling relative slow evolution[@Patel2018].

I use the same principle that Patel et al. use to modulate the relative rate of evolution to modulate rates of component responses for $S$ components. Following May[@May1972; @May1973], the value of a component $i$ at time $t$ ($v_{i}(t)$) is affected by the value of $j$ ($v_{j}(t)$) and $j$'s marginal effect on $i$ ($a_{ij}$), and by $i$'s response rate ($\gamma_{i}$),

In the above, $\mathbf{\gamma}$ is a diagonal matrix in which elements correspond to individual component response rates. Therefore, $\mathbf{M} =\mathbf{\gamma} \mathbf{A}$ modulates the values of components and can be analysed using the techniques of May[@May1972; @May1973; @Ahmadian2015].

**Genetic algorithm**. Ideally, to investigate the potential of $Var(\gamma)$ for increasing the proportion of stable complex systems, the search space of all possible $\gamma$ vectors would be evaluated for each unique $\mathbf{M = \gamma A}$. This is technically impossible because $\gamma_{i}$ can take any real value between 0-2, but even rounding $\gamma_{i}$ to reasonable values would result in a search space too large to practically explore. Under these conditions, genetic algorithms are highly useful tools for finding practical solutions by mimicking the process of biological evolution[@Hamblin2013]. In this case, the practical solution is finding vectors of $\mathbf{\gamma}$ that decrease the most positive real eigenvalue of $\mathbf{M}$. The genetic algorithm used achieves this by initialising a large population of 1000 different potential $\mathbf{\gamma}$ vectors and allowing this population to evolve through a process of mutation, crossover (swaping $\gamma_{i}$ values between vectors), selection, and reproduction until either a $\mathbf{\gamma}$ vector is found where all $\Re(\lambda) < 0$ or some "giving up" critiera is met.

**System feasibility**. Dougoud et al.[@Dougoud2018] define the following feasibility criteria for ecological systems characterised by $S$ interacting species with varying densities in a classical Lotka-Volterra model,

In the above, $\mathbf{x^{*}}$ is the vector of species densities at equilibrium. Feasibility is satisfied if all elements in $\mathbf{x^{*}}$ are positive. The matrix $\mathbf{I}$ is the identity matrix, and the value $\theta$ is the strength of intraspecific competition (diagonal elements). Diagonal values are set to $-1$, so $\theta = -1$. The variable $\delta$ is a normalisation parameter that modulates the strength of interactions ($\sigma$) for $\mathbf{J}$. Implicitly, here $\delta = 0$ underlying strong interactions. Hence, $(CS)^{-\delta} = 1$, so in the above, a diagonal matrix of -1s ($\theta \mathbf{I}$) is added to $\mathbf{J}$, which has a diagonal of all zeros and an off-diagonal affecting species interactions (i.e., the expression $(CS)^{-\delta}$ relates to May's[@May1972] stability criterion[@Dougoud2018] by $\frac{\sigma}{(CS)^{-\delta}}\sqrt{SC} < 1$, and hence for my purposes $(CS)^{-\delta} = 1$). Given $\mathbf{A} = \theta\mathbf{I + J}$, the above criteria is therefore reduced to the below,

>>>>

---
title: "Component response rate variation drives stability in large complex systems"
author: "A. Bradley Duthie    ( alexander.duthie@stir.ac.uk )"
date: Biological and Environmental Sciences, University of Stirling, Stirling, UK,
  FK9 4LA
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
  html_document: default
  word_document:
    fig_caption: yes
    pandoc_args:
    - --csl
    - science.csl
    reference_docx: docx_template.docx
bibliography: references.bib
header-includes:
- \usepackage{amsmath}
- \usepackage{natbib}
- \usepackage{lineno}
- \usepackage[utf8]{inputenc}
- \linenumbers
- \bibliographystyle{amnatnat}
linestretch: 2
link-citations: no
linkcolor: blue
csl: science.csl
biblio-style: apalike
---

@@

Higher stability in systems with variation in $\gamma$ can be observed by sampling $\gamma_{i}$ values from various distributions. I focus on a uniform distribution where $\gamma \sim \mathcal{U}(0, 2)$ (see Supplementary Information for other distributions, which give similar results). As with the case of $\gamma = \{1.95, 0.5\}$ (Fig. 1b), $E[\gamma] = 1$ when $\gamma \sim \mathcal{U}(0, 2)$, allowing comparison of $\mathbf{M}$ before and after the addition of variation in component response rate. Figure 2 shows a comparison of eigenvalue distributions given $S = 1000$, $C = 0.05$, and $\sigma = 0.4$. As expected [@Tao2010], when $\gamma = 1$, eigenvalues are distributed uniformly in a circle centred at ($-1, 0$) with a radius of $\sigma\sqrt{SC} =$ `r round(sigma*sqrt(S*C), digits = 3)`. Uniform variation in $\gamma$ leads to a non-uniform distribution of eigenvalues [@Ahmadian2015; @Gibbs2017; @Stone2017], some of which are clustered locally near the centre of the distribution, but others of which are spread outside the former radius of `r round(sigma*sqrt(S*C), digits = 3)` (Fig 2b). The clustering and spreading of eigenvalues introduced by $Var(\gamma)$ can destabilise previously stable systems or stabilise systems that are otherwise unstable. But where systems are otherwise too complex to be stable given $\gamma = 1$, the effect of $Var(\gamma)$ can often lead to stability above May's [@May1972; @Allesina2012] threshold $\sigma\sqrt{SC} < 1$.

@@

**Targeted manipulation of $\mathbf{\gamma}$**. To further investigate the potential of $Var(\gamma)$ to be stabilising, I used a genetic algorithm. Genetic algorithms are heuristic tools that mimic evolution by natural selection, and are useful when the space of potential solutions (in this case, possible combinations of $\gamma$ values leading to stability in a large complex system) is too large to search exhaustively [@Hamblin2013]. Generations of selection on $\gamma$ value combinations to minimise $\max\left(\Re(\lambda)\right)$ demonstrated the potential for $Var(\gamma)$ to increase system stability. Across $S = \{2, 3, ..., 39, 40\}$, sets of $\gamma$ values were found that resulted in stable systems with probabilities that were up to four orders of magnitude higher than when $\gamma = 1$ (Fig. 4), meaning that stability could often be achieved by manipulating $S$ $\gamma$ values rather than $S \times S$ $\mathbf{M}$ elements.

**System feasibility given $\mathbf{Var(\gamma)}$** For complex systems in which individual system components represent the density of some tangible quantity, it is relevant to consider the feasibility of the system. Feasibilility assumes that the system is stable and values of all components are positive at equilibrium [@Grilli2017; @Dougoud2018; @Song2018]. This is of particular interest for ecological communities because population density cannot take negative values, meaning that ecological systems need to be feasible for stability to be biologically realistic [@Dougoud2018]. While my results are intended to be general to all complex systems, and not restricted to species networks, I have also performed a feasibility analysis on all matrices $\mathbf{M}$ tested for stability, and additionally for specific types of ecological communities [@Allesina2012] (e.g., competitive, mutualist, predator-prey; see Supplementary Information). Feasibility was unaffected by $Var(\gamma)$, meaning that for pure interacting species networks, variation in component response rate (i.e., species generation time) does not affect stability at biologically realistic species densities.

Here I have shown that the stability of large systems might often be contigent upon variation in the response rates of their individual components, meaning that factors such as rate of trait evolution (in biological networks), transaction speed (in economic networks), or communication speed (in social networks) need to be considered when investigating the stability of complex systems. Variation in component response rate becomes more likely to be critical for stability as system size increases, and can ultimately increase the overall probability that system stability is observed above that predicted by May's [@May1972] classically derived $\sigma \sqrt{SC}$ criterion. The logic outlined here is general, and potentially applies to any complex system in which individual system components can vary in their reaction rates to system perturbation.

It is important to emphasise that variation in component response rate is not stabilising per se; that is, adding variation in component response rate to a particular system does not necessarily increase the probability that the system will be stable. Rather, systems that are observed to be stable are more likely to have varying component response rates, and for this variation to be critical to their stability (Fig. 3). This is caused by the shift to a non-uniform distribution of eigenvalues that occurs by introducing $Var(\gamma)$ (Fig. 1b, 2b), which can sometimes cause all of the real components of the eigenvalues of the community matrix $\mathbf{M}$ to become negative, but might also increase the real components of eigenvalues. The mathematics underlying this shift in eigenvalue distribution has been investigated [@Ahmadian2015] and recently applied to questions concerning species density and feasibility [@Gibbs2017; @Stone2017], but has not been interpreted as rates of response of individual system components to perturbation.

Interestingly, while complex systems were more likely to be stable given variation in component response rate, they were not more likely to be feasible, meaning that stability was not increased when component values were also restricted to being positive at equilibrium. Feasibility is important to consider, particularly for the study of ecological networks of species [@Stone2017; @Grilli2017; @Dougoud2018] because population densities cannot realistically be negative. My results therefore suggest that variation in the rate of population responses to perturbation (e.g., due to differences in generation time among species) is unlikely to be critical to the stability of purely multi-species interaction networks (see also Supplementary Information). Nevertheless, ecological interactions do not exist in isolation in empirical systems [@Patel2018], but instead interact with evolutionary, abiotic, or social-economic systems. The relevance of component response rate for complex system stability should therefore not be ignored in the broader context of ecological communities.

In the above, $f_{i}$ and $g_{j}$ are functions that define the effects of all species densities and trait values on the density of a species $i$ and the value of trait $j$, respectively. Patel et al. were interested in stability when the evolution of traits was relatively slow or fast in comparison with the change in species densities [@Patel2018], and this is modulated in the above by the scalar $\epsilon$. The value of $\epsilon$ thereby determines the timescale separation between ecology and evolution, with high $\epsilon$ modelling relatively fast evolution and low $\epsilon$ modelling relative slow evolution [@Patel2018].

I use the same principle that Patel et al. use to modulate the relative rate of evolution to modulate rates of component responses for $S$ components. Following May [@May1972; @May1973], the value of a component $i$ at time $t$ ($v_{i}(t)$) is affected by the value of $j$ ($v_{j}(t)$) and $j$'s marginal effect on $i$ ($a_{ij}$), and by $i$'s response rate ($\gamma_{i}$),

In the above, $\mathbf{\gamma}$ is a diagonal matrix in which elements correspond to individual component response rates. Therefore, $\mathbf{M} =\mathbf{\gamma} \mathbf{A}$ modulates the values of components and can be analysed using the techniques of May [@May1972; @May1973; @Ahmadian2015].

**Genetic algorithm**. Ideally, to investigate the potential of $Var(\gamma)$ for increasing the proportion of stable complex systems, the search space of all possible $\gamma$ vectors would be evaluated for each unique $\mathbf{M = \gamma A}$. This is technically impossible because $\gamma_{i}$ can take any real value between 0-2, but even rounding $\gamma_{i}$ to reasonable values would result in a search space too large to practically explore. Under these conditions, genetic algorithms are highly useful tools for finding practical solutions by mimicking the process of biological evolution [@Hamblin2013]. In this case, the practical solution is finding vectors of $\mathbf{\gamma}$ that decrease the most positive real eigenvalue of $\mathbf{M}$. The genetic algorithm used achieves this by initialising a large population of 1000 different potential $\mathbf{\gamma}$ vectors and allowing this population to evolve through a process of mutation, crossover (swaping $\gamma_{i}$ values between vectors), selection, and reproduction until either a $\mathbf{\gamma}$ vector is found where all $\Re(\lambda) < 0$ or some "giving up" critiera is met.

**System feasibility**. Dougoud et al. [@Dougoud2018] define the following feasibility criteria for ecological systems characterised by $S$ interacting species with varying densities in a classical Lotka-Volterra model,

In the above, $\mathbf{x^{*}}$ is the vector of species densities at equilibrium. Feasibility is satisfied if all elements in $\mathbf{x^{*}}$ are positive. The matrix $\mathbf{I}$ is the identity matrix, and the value $\theta$ is the strength of intraspecific competition (diagonal elements). Diagonal values are set to $-1$, so $\theta = -1$. The variable $\delta$ is a normalisation parameter that modulates the strength of interactions ($\sigma$) for $\mathbf{J}$. Implicitly, here $\delta = 0$ underlying strong interactions. Hence, $(CS)^{-\delta} = 1$, so in the above, a diagonal matrix of -1s ($\theta \mathbf{I}$) is added to $\mathbf{J}$, which has a diagonal of all zeros and an off-diagonal affecting species interactions (i.e., the expression $(CS)^{-\delta}$ relates to May's[@May1972] stability criterion [@Dougoud2018] by $\frac{\sigma}{(CS)^{-\delta}}\sqrt{SC} < 1$, and hence for my purposes $(CS)^{-\delta} = 1$). Given $\mathbf{A} = \theta\mathbf{I + J}$, the above criteria is therefore reduced to the below,

\clearpage

@@

\clearpage

@@

\clearpage

@@

\clearpage

@@
